Staged Abstract Interpreters

### 2019-09-06 paper/30_stageconc.tex
typo

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show that how to stage the concrete
interpreter by changing the monad type and refactoring primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala, and then replay the same steps from the unstaged
counterpart. Finally, we briefly describe code generation.

@@

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} which
use syntactic quotations, LMS distinguishes binding-time based on types. 
LMS provides a type constructor @Rep[T]@ where @T@ can be
an arbitrary type, indicating a value of type @T@ will be known at future stage.
Operations acting on @Rep[T]@ expressions will be residualized as generated
code.

A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

If @x@ is a value known at the current stage, we may specialize the power function to
the value @x@ -- by unrolling the recursive calls. In LMS, this is fulfilled by
first adding the @Rep@ type annotation to variables known at next stage. In this case,
@b : Rep[Int]@ is known later, and @x : Int@ is known currently, as shown in the
below code recipe.
The way we use this staged @power@ function is to create a @DslDriver@ and override the
@snippet@ method that gives the currently know value @x@ (5 in this example).
\begin{lstlisting}
  new DslDriver[Int, Int] {
    def power(b: Rep[Int], x: Int): Rep[Int] = if (x == 0) 1 else b * power(b, x-1)
    def snippet(b: Rep[Int]): Rep[Int] = power(b, 5) // specialize the power to b^5
  }
\end{lstlisting}

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as lists and maps.
The idea behind the framework is to construct a sea-of-node intermediate
representations (IR) for the next-stage program at the current stage
\cite{DBLP:conf/birthday/Rompf16}. For convenience, the conversion from
expressions of type @Rep[_]@ to their IR is done by using implicit design
pattern. As we will see later, implementing staging and code generation
support for user-defined classes is also straightforward.

@@

To implement the staged concrete interpreter, we replay the steps from the
instantiation of unstaged concrete interpreter in Section \ref{unstaged_conc}.
But, now we use the @Rep@ type to annotate the value domains, environments and
stores, and redefine the staged version of monads and primitive operations.

\paragraph{Staged Monads}
We use the same structure of monad stack in the unstaged interpreter: a reader
monad with a state monad. The difference here is, now the monads operate on staged values. 
For brevity, we call them \textit{staged monads}. In the code, we also use a @Rep@
prefix on the constructors and types to differentiate them. But it is important
to note that instances of the case class @ReaderT@/@StateT@ are not staged, also the
monadic computation like @R => M[A]@ are also not staged. Instead, the internal
data that these monads operate on are staged, i.e., @R@ and @A@ in the
reader monad, @S@ and @A@ in the state monad. The following code
snippet shows the idea. We use \hl{light gray} to highlight where @Rep@ type is added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a function value known at current stage.
Therefore all invocations of @flatMap@s can be reduced before generating code.
The fact that we only stage data but not monadic computation or monadic
values, is the reason that we can peel of the monad stack in the generated code.
Now type @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions:
\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not straightforward for nondeterminism monad
(@SetT@) in abstract interpreters. Because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements in the set. In this section, we explicitly distinguish the two
versions of monadic interfaces, later we will instantiated the staged set monad
by manually fusing part of the monad stack.

>>>>

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete
interpreter by changing the monad type and refactoring primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala, and then replay the same steps from the unstaged
counterpart. Finally, we briefly describe the code generation.

@@

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} which
use syntactic quotations, LMS distinguishes binding-time based on types. 
LMS provides a type constructor @Rep[T]@ where @T@ can be
an arbitrary type, indicating a value of type @T@ will be known at the future stage.
Operations acting on @Rep[T]@ expressions will be residualized as generated
code.

A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

If @x@ is a value known at the current stage, we may specialize the power function to
the value @x@ -- by unrolling the recursive calls. In LMS, this is fulfilled by
first adding the @Rep@ type annotation to variables known at the next stage. In this case,
@b : Rep[Int]@ is known later, and @x : Int@ is known currently, as shown in the
below code recipe.
The way we use this staged @power@ function is to create a @DslDriver@ and override the
@snippet@ method that gives the currently know value @x@ (5 in this example).
\begin{lstlisting}
  new DslDriver[Int, Int] {
    def power(b: Rep[Int], x: Int): Rep[Int] = if (x == 0) 1 else b * power(b, x-1)
    def snippet(b: Rep[Int]): Rep[Int] = power(b, 5) // specialize the power to b^5
  }
\end{lstlisting}

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as lists and maps.
The idea behind the framework is to construct a sea-of-node intermediate
representation (IR) for the next-stage program at the current stage
\cite{DBLP:conf/birthday/Rompf16}. For convenience, the conversion from
expressions of type @Rep[_]@ to their IR is done by using the implicit design
pattern. As we will see later, implementing staging and code generation
support for user-defined classes is also straightforward.

@@

To implement the staged concrete interpreter, we replay the steps from the
instantiation of the unstaged concrete interpreter in Section \ref{unstaged_conc}.
However, now we use the @Rep@ type to annotate the value domains, environments and
stores, and redefine the staged version of monads and primitive operations.

\paragraph{Staged Monads}
We use the same structure of monad stack in the unstaged interpreter: a reader
monad with a state monad. The difference here is, now the monads operate on
staged values.  For brevity, we call them \textit{staged monads}. In the code,
we also use a @Rep@ prefix on the constructors and types to differentiate them.
But it is important to note that instances of the case class @ReaderT@/@StateT@
are not staged; also the monadic computation, such as functions of @R => M[A]@,
are not staged. Instead, the internal data that these monads operate on are
staged, i.e., @R@ and @A@ in the reader monad, @S@ and @A@ in the state monad.
The following code snippet shows the idea. We use \hl{light gray} to highlight
where @Rep@ type is added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a function value known at the current stage.
Therefore all invocations of @flatMap@s can be reduced before generating code.
The fact that we only stage data but not monadic computation or monadic
values is the reason that we can peel of the monad stack in the generated code.
Now the type @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions:
\begin{lstlisting}
  type R[T] = Rep[T]
  type AnsM[T] = RepReaderT[RepStateT[RepIdM, Store, ?], Env, T]
\end{lstlisting}

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not straightforward for nondeterminism monad
(@SetT@) in the abstract interpreter. Because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements in the set. In this section, we explicitly distinguish the two
versions of monadic interfaces; later we will instantiate the staged set monad
by manually fusing a fragment of the monad stack.
